Domain Specific Languages and Code Synthesis using Haskell

### 2014-03-14 paper.tex
Rest of paper edits

The basis of a Deep EDSL is one of constructiveness.
Functional programming is about constructing {\em and destructing\/} values.
Because of this, a deep embedding can not reify any pattern matching, 
or even direct usage of if-then-else, and other control flow. We side-stepped this
in Kansas Lava, for example by using a \verb|mux2| constructor, 
which encodes choice.
So, how much further can the idiom be pushed, if we need to be constructive?
The result is surprising. We start with our three capabilities:
\begin{itemize}
\item Basic expressions can be captured, by constructing a tree that is an analog to our syntax.
\item Functions can be captured, using the a fake unique argument.
\item Local bindings can be observed using some form of observable sharing.
\end{itemize}
With these three comes an automatic forth capability.
\begin{itemize}
\item The host language provides a built-in macro capability to the embedded language.
Any part of Haskell (including control flow and pattern matching) can be used
to {\em generate\/} the embedded language. 
\end{itemize}

\noindent
There are also extensions to the basic techniques. The principal ones are:
\begin{itemize}
\item Internal function calls can captured as notes on a graph, 
rather than directly inlined~\cite{Mainland:10:Nikola}. 
This helps compilation of large programs, giving a basic separate compilation capability.
\item \verb|do|-statement can be reified, by normalization~\cite{Sculthorpe:13:ConstrainedMonad,Svenningsson:13:CompositionalMonadReification,Persson:11:GenericMonadicConstructs}. This result, called monadic
reification, is surprising. There are strong technical reasons to be believe monadic reification
should be impossible. However the normalization refactors the constrains that, by themselves would
be impossible to solve, and matches them up, 1-on-1, with a matching witness, allowing the whole
\verb|do|-notation to be solved and reified.
Monadic reification is a recent discovery, but has already been used in several Deep DSLs,
including Feldspar\cite{Axelsson:2010:DandI-Feldspar} and Sunroof\cite{Bracker:14:Sunroof}.
\item Control flow is problematic, and can not be used directly. However,
there is a generalization of Haskell boolean that does allow deep embedding capture~\cite{hackage:boolean}.
Using this library, a DSL with control flow can be constructed, but it needs to be explicit
code, at the DSL level, using constructors. The usage is clumsy, but workable, and we should
be able to do better.
\end{itemize}

@@

Parametric polymorphism, one of the strengths of functional program, is the other
issue for deep DSLs. We need a specific structure to represent what we have captured,
and arbitrary polymorphism interferes with this. Current system sidestep this
by always instantiate at a specific type, but this expensive because size of
the captured program can expand exponentially.
Polymorphism was the technical reason that it was thought that monadic reification
would not work, but in that case it was side-stepped by normalization; this technique
does not generalize to all polymorphism.

Deep DSLs is the value-level way of extracting an expression, but there are others.
Quaziquoting is another, at the syntactically level. Haskell comes with an extensive
template system, called Template Haskell, that can and is used for DSLs.
Perhaps the future of Deep DSLs is some hybrid between expression generation
and quasi-quoting, combining the best of both systems?

>>>>

The basis of a Deep EDSL is one of constructiveness.
Functional programming is about constructing {\em and destructing\/} values.
Because of this, a deep embedding cannot reify any pattern matching, 
or even direct usage of if-then-else, and other control flow. We side-stepped this
in Kansas Lava, for example by using a \verb|mux2| constructor, 
which encodes choice.
So, how much further can the idiom be pushed, if we need to be constructive?
The result is surprising. We start with our three capabilities:
\begin{itemize}
\item Basic expressions can be captured, by constructing a tree that is an analog to our syntax.
\item Functions can be captured, using a fake unique argument.
\item Local bindings can be observed using some form of observable sharing.
\end{itemize}
With these three comes an automatic fourth capability.
\begin{itemize}
\item The host language provides a built-in macro capability to the embedded language.
Any part of Haskell (including control flow and pattern matching) can be used
to {\em generate\/} the embedded language. 
\end{itemize}

\noindent
There are also extensions to the basic techniques. The principal ones are:
\begin{itemize}
\item Internal function calls can captured as notes on a graph, 
rather than directly inlined~\cite{Mainland:10:Nikola}. 
This helps compilation of large programs, giving a basic separate compilation capability.
\item \verb|do|-statement can be reified, by normalization~\cite{Sculthorpe:13:ConstrainedMonad,Svenningsson:13:CompositionalMonadReification,Persson:11:GenericMonadicConstructs}. This result, called monadic
reification, is surprising. There are strong technical reasons to be believe monadic reification
should be impossible. However the normalization refactors the constraints that, by themselves, would
be impossible to solve, and matches them up, 1-on-1, with a matching witness, allowing the whole
\verb|do|-notation to be solved and reified.
Monadic reification is a recent discovery, but has already been used in several Deep DSLs,
including Feldspar\cite{Axelsson:2010:DandI-Feldspar} and Sunroof\cite{Bracker:14:Sunroof}.
\item Control flow is problematic, and cannot be used directly. However,
there is a generalization of Haskell boolean that does allow deep embedding capture~\cite{hackage:boolean}.
Using this library, a DSL with control flow can be constructed, but it needs to be explicit
code, at the DSL level, using constructors. The usage is clumsy, but workable, and we should
be able to do better.
\end{itemize}

@@

Parametric polymorphism, one of the strengths of functional program, is the other
issue for deep DSLs. We need a specific structure to represent what we have captured,
and arbitrary polymorphism interferes with this. Current systems sidestep this
by always instantiating at a specific type, but this is expensive because the size of
the captured program can expand exponentially.
Polymorphism was the technical reason that it was thought that monadic reification
would not work, but in that case it was side-stepped by normalization; this technique
does not generalize to all polymorphism.

Deep DSLs is a value-level way of extracting an expression, but there are others.
Quasi-quoting is another, at the syntactic level. Haskell comes with an extensive
template system, called Template Haskell, that is used for DSLs.
Perhaps the future of Deep DSLs is some hybrid between expression generation
and quasi-quoting, combining the best of both systems?
